Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-23 isolation_by_coalescence.tex
edits from call

\newcommand{\comdist}{\widetilde R}

@@

The idea of ``resistance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin}.
Both electrical resistance and commute time between two nodes
average over all possible paths through the network between the two.
For this reason, it is thought that if genetic distances between samples
from different parts of the landscape
are significantly correlated with the effective resistances between those locations,
with local conductances assigned based on the values of some landscape variable,
then that landscape variable is likely a good indicator of the regions of the map 
where gene flow occurs.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is set equal to its elevation,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more at higher elevations,
perhaps along ridgelines.
\plr{reference some examples}
Other methods \citep{eems} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

Genetic distances between two genomes sampled from the landscape
do derive from an average across a large number of paths between those points --
the lineages along which each segment of genetic material has been inherited
from their most recent common ancestor.
The more recent this ancestor is,
the smaller the genetic distance will be.
Thanks to recombination, there are a large number of such paths, 
and genetic distance averages over these.
It is reasonable in some situations to model these lineages as random walks across the landscape;
however, the model of genetic distance that one is thus led to
is \emph{coalescence time} of the random walks, 
rather than commute time, the quantity that corresponds to resistance.
This naturally raises the question: 
Are methods that depend on effective resistances being misled?
If so, would using coalescence time do better?
To better contrast the two,
we refer to ``commute time'' rather than ``resistance'' throughout,
although the two concepts are equivalent.

@@

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the mean time it takes a particle moving according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the mean time it takes two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
First, we explain this connection,
and then discuss their computation and application.

@@

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, we will find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so we must account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing another set of model parameters, $q$, 
that fill the same role as the coalescence rates, $\gamma$.
Then, we use commute time to approximate genetic distance as
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
The factor of four appears to make commute time more similar to coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
Furthermore, coalescence time includes within-location diversity
because particles do not coalesce instantly --
therefore commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

@@

Although commute times are uniquely determined by movement rates, the reverse is not true.
Commute times only depend on the symmetric part of $H$ --
given an antisymmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$
(assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
(It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.
\plr{say a bit more about how we solve this?}

@@

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned} 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
produce large differences in the inferred rates.
Very roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
such operators are deformations of the Laplacian \citep{laplacian},
which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.

@@

Consider a Markov chain with two states, 
where the rate of movement from state 1 to state 2 is $G_{12}$
and the rate of movement from state 2 to state 1 is $G_{21}$,
so the generator matrix is
\begin{align*}
G = 
    \begin{bmatrix}
        -G_{12}  & G_{12} \\
         G_{21}  & -G_{21}
    \end{bmatrix}.
\end{align*}
Writing out equation \ref{eqn:C_matrix}, we get three unique equations
for the coalescence times:
\plr{simplify these equations and remove matrix version?}
\begin{align*}
   -C_{11} G_{12} + C_{21} G_{12} - C_{11} G_{12} + C_{12} G_{12} - C_{11} \gamma_1 &= -1 \\
    C_{11} G_{21} - C_{21} G_{21} - C_{21} G_{12} + C_{22} G_{12}                   &= -1 \\
    C_{12} G_{21} - C_{22} G_{21} + C_{21} G_{21} - C_{22} G_{21} - C_{22} \gamma_2 &= -1 .
\end{align*}
Given $C$, we have three equations for the four unknowns, 
$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$.
By grouping terms together, factoring back into matrix form, 
and writing $C_{ji}$ instead of $C_{ij}$ when $i > j$,
we have
\begin{align*}
\begin{bmatrix}
-2C_{11} + 2C_{12} & 0                 & -C_{11} & 0       \\
-C_{12} + C_{22}   & C_{11} - C_{12}   & 0       & 0       \\
0                  & 2C_{12} - 2C_{22} & 0       & -C_{22}
\end{bmatrix}
\begin{bmatrix}
G_{12} \\ G_{21} \\ \gamma(1) \\ \gamma(2)
\end{bmatrix}
=
\begin{bmatrix}
-1 \\ -1 \\ -1
\end{bmatrix}.
\end{align*}
This is small enough to be solved symbolically. 
The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
and $\gamma_2$ nonnegative.
If $C_{11} \neq C_{12}$, 
we can write the solution with $\gamma_1$ as the free variable:
\begin{align*}
G_{12} &= \frac{1}{2(C_{11} - C_{12})} - \frac{C_{11}}{2(C_{11} - C_{12})}\gamma_1 \\
G_{21} &= \frac{-2(C_{11} - C_{12}) + (C_{12} - C_{22})}{2(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})}{2(C_{11} - C_{12})^2}\gamma_1 \\
\gamma_2 &= \frac{(C_{11} - 2C_{12} + C_{22})^2}{C_{22}(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})^2}{C_{22}(C_{11} - C_{12})^2}\gamma_1.
\end{align*}
This implies that $\gamma_2$ decreases as $\gamma_1$ increases as long as $C_{12} \neq C_{22}$.
This makes sense because
increasing a rate of coalescence cannot make expected coalescence times longer,
so in order to keep coalescence times the same,
if one coalescence rate is increased, another must be decreased.
This also implies that, in order for $\gamma_2$ to be a value other than $0$, 
we must have $C_{11} - 2C_{12} + C_{22} \neq 0$.
In general, if there is a value or range of values of $\gamma_1$
for which the other movement and coalescence parameters are non-negative,
then a solution exists.

@@

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
the movement rates, $G_{ij}$, that are not constrained to be zero
and the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph,
and only movement rates corresponding to edges in the graph are allowed to be nonzero.
Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
that there are $m_G$ edges in the graph,
and that each of the $n$ populations has its own coalescence rate.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.

Inference with commute times is done in exactly the same way,
except that $q$ replaces $\gamma$,
and instead of $\mathcal{C}(G,\gamma)$ we use $\comdist(G, q)$
as defined in equation \eqref{eq:commute_approx}.

@@

We compared inference under coalescence and commute time models
using both data ``from the model'' and more realistic simulations.
For the first category, we designed a number of graphs to test particular aspects of inference,
with migration rates specified on each directed edge
(plotted below using igraph \citep{igraph}).
Then we calculated exact coalescence and commute times on these graphs,
and added Gaussian noise independently to each entry of the resulting matrices
(while preserving symmetry)
as described above.
Particular parameters and levels of noise are given in the Results.

@@

We first tested the method using data generated under the model.
To test the methods across a wide range of situations,
we generated random movement rates for 
a large number of $3 \times 2$ rectangular grids:
twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
We generated movement parameters
by rounding independent exponential random variables up to the nearest one tenth,
and fixed coalescence rate at 1 for all locations.
We then added independent noise to each matrix entry (preserving symmetry),
and assume that we can estimate the amount of noise reasonably well, 
so used the standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
To assess the impact on inference of the number of parameters,
we performed inference both with and without the assumption 
that coalescence rate is the same everywhere.
For $3 \times 2$ graphs, there are 21 equations
and 14 movement parameters,
so that with a global coalescence rate there are 15 unknowns
but with individual coalescence rates there are 20 unknowns,
which is close to nonidentifiable.
(For commute time inference, we analogously constrained all values of $q$ to be identical or not.)
To quantify accuracy across these random graphs in different situations,
we computed, for each model fit on a particular random graph,
the absolute difference between the posterior median of each movement parameter
and the true value.
We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
reported below as ``mean absolute error''.

@@

We then test the method's ability to locate barriers to migration
on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
Each nonzero movement rate is determined randomly as before.
Coalescence rates are set to 1 everywhere.
Figure \ref{fig:5x3boxplots} shows posterior distributions of the movement rates
for both coalescence and commute time inference
on this graph, with noise standard deviation equal to
1/1000 of the mean coalescence time (left)
and 1/100 of the mean coalescence time (right).
Although increased noise in the data increased uncertainty in the estimated values,
coalescence time inference correctly inferred the location of the barrier
(as well as high-dispersal connections) in both cases.
Commute time inference \plr{did what?}
We also explored making the problem harder,
by first (a) dropping rows and columns from the data corresponding to populations 2, 5, 7, and 15;
and then (b) allowing a separate coalescence rate for each location,
in both cases using a noise level of 1/100.
The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
and show substantially increased error,
but only in movement rates to the removed locations.
As before, uncertainty was increased when multiple coalescence rates were allowed,
but not as strongly as before, likely because the problem is less close to being underconstrained,
having 105 equations and 59 unknowns.

\begin{figure}
\centering
     \includegraphics[scale=.6]{figs/5x3b_grid}
    \caption{Plot of the grid structure 
    for the 5 $\times$ 3 grid with two barriers:
    movement rates between $\{2,3\}$ and $\{7,8\}$ are zero,
    as are those between $\{8,9\}$ and $\{13,14\}$.
    Values for the non-zero movement rates 
    were chosen by rounding independent draws of an exponential random variable with mean 1 
    up to the nearest tenth.}
    \label{fig:5x3b_grid}
\end{figure}

\begin{figure}
\centering
     \includegraphics[scale=0.8]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
    \plr{TODO: replace bottom two panels with same thing for commute time}
}
    \label{fig:5x3boxplots}
\end{figure}

@@

Both matrices of coalescence and commute times, $C$ and $R$, are symmetric,
but they do not deal with asymmetric (biased) migration in the same way.
Since commute time to $i$ requires paths both to and from $i$,
but coalescence time to $i$ only reaquires lineages to leave $i$,
if, for instance, there is a low rate of migration back into $i$,
then commute times to $i$ could be quite long even while coalescence times are not.
For instance, suppose there are three populations arranged in a line,
and lineages currently in the outer populations are much more likely to come from the central island
than the other way around.
Since lineages quickly move to the center and coalesce regardless of starting position,
coalescence time (and genetic diversity) will be relatively low between all individuals.
Commute time between the two outer locations, on the other hand,
will be much longer than other comparisons, since it requires a lineage to leave the center.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/4x4coalvcomH}
    \caption{Plots of the true values of $\comdist$ vs $H$ 
    for the uniform graph, symmetric graph, the asymmetric graph,
    and the biased asymmetric graph, 
    shown with the $y=x$ line.
    \plr{TODO: make this a 2x2 grid instead of 4x1}
    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
    \label{fig:4x4coalvcomH}
\end{figure}

>>>>

\newcommand{\comdist}{\widetilde R}
\newcommand{\comdistfn}{\widetilde \mathcal{R}}

@@

The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin}.
Both electrical resistance and commute time between two nodes
average over all possible paths through the network between the two.
For this reason, it is thought that if genetic distances between samples
from different parts of the landscape
are significantly correlated with the effective resistances between those locations,
with local conductances assigned based on the values of some landscape variable,
then that landscape variable is likely a good indicator of the regions of the map 
where gene flow occurs.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is higher in flatter areas,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more readily through flatter areas,
perhaps along river valleys.
\plr{reference some examples}
Other methods \citep{eems} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

Genetic distances between two genomes sampled from the landscape
do derive from an average across a large number of paths between those points --
the lineages along which each segment of genetic material has been inherited
from their most recent common ancestor.
The more recent this ancestor is,
the smaller the genetic distance will be.
Thanks to recombination, there are a large number of such paths, 
and genetic distance averages over these.
It is reasonable in some situations to model these lineages as random walks across the landscape;
however, the model of genetic distance that one is thus led to
is \emph{coalescence time} of the random walks, 
rather than commute time, the quantity that corresponds to resistance.
This naturally raises the question: 
Are methods that depend on effective resistances being misled?
If so, would using coalescence time do better?
To better contrast the two,
we refer to ``commute time'' rather than ``resistance'' throughout,
although the two concepts are equivalent.
\plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}

@@

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the expected time for a particle that moves according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the expected time for two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
First, we explain this connection,
and then discuss their computation and application.

@@

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, we will find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so we must account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing another set of model parameters, $q$, 
that fill an analogous role to the coalescence rates, $\gamma$.
Then, we use commute time to approximate genetic distance as
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
The factor of four appears to make commute time more similar to coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
Furthermore, coalescence time includes within-location diversity
because particles do not coalesce instantly --
therefore, commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

@@

Although commute times are uniquely determined by movement rates, the reverse is not true.
Commute times only depend on the symmetric part of $H$ --
given an skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$
(assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
(It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

% \paragraph{Coalescence times}
The \emph{coalescence times} are defined using \emph{two} copies of the same chain,
as the mean time until coalescence,
if the chains coalesce at rate $\gamma$ when they are in the same place.
Concretely, 
suppose that $X$ and $Y$ are independent Markov chains both moving with movement rates given by $G$,
that coalesce at rate $\gamma_x$ when $X$ and $Y$ are both at $x$.
Then we define $\tau$ to be this coalescence time,
so that 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t = Y_t = x\} = \epsilon \gamma_x + O(\epsilon^2)$,
and 
$\P\{\tau \le t + \epsilon \given \tau > t, \; X_t \neq Y_t\} = O(\epsilon^2)$.
Then, the (mean) coalescence time is $C_{xy} = \E[\tau \given X_0 = x, \, Y_0 = y]$.
A similar argument using the Markov property
shows that $C$ satisfies the following equation:
\begin{align}
\text{\bf (coalescence time)} \qquad
    \sum_y \left(G_{xy} C_{yz} + C_{xy} G_{zy}\right)
    &=
    \begin{cases}
        -1                   \quad & \text{if} \; x \neq z, \\
        -1 + \gamma_z C_{zz} \quad & \text{if} \; x = z.
    \end{cases}
\end{align}
In matrix notation, this is
\begin{align} \label{eqn:C_matrix}
    G C + C G^T - \diag(\gamma) \circ C = -\bone \bone^T,
\end{align}
where $\diag(\gamma)$ is the matrix with the vector $\gamma$ on the diagonal and zeros elsewhere,
and $\circ$ is the componentwise product.
In practice, we solve this by working with the product Markov chain $(X_t, Y_t)$,
whose generator matrix is $G \otimes I + I \otimes G$,
where $I$ is the identity matrix and $\otimes$ is the Kronecker product.

@@

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned} 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
may produce large differences in the inferred rates.
Very roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
such operators are deformations of the Laplacian \citep{laplacian},
which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.

@@

Consider a Markov chain with two states, 
where the rate of movement from state 1 to state 2 is $G_{12}$
and the rate of movement from state 2 to state 1 is $G_{21}$,
so the generator matrix is
\begin{align*}
G = 
    \begin{bmatrix}
        -G_{12}  & G_{12} \\
         G_{21}  & -G_{21}
    \end{bmatrix}.
\end{align*}
Equation \ref{eqn:C_matrix} equates two $2 \times 2$ matrices,
so provides four equations. Only three of these are unique;
simplifying these and using that $C_{12} = C_{21}$, these are:
\begin{align*}
    2 \left(C_{12} - C_{11} \right) G_{12} - C_{11} \gamma_1 &= -1 \\
    \left(C_{11} - C_{12} \right) G_{21} + \left(C_{22} - C_{12}\right) G_{12} &= -1 \\
    2 \left( C_{12} - C_{22} \right) G_{21} - C_{22} \gamma_2 &= -1 .
\end{align*}
Given $C$, we have three equations for the four unknowns, 
$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$,
which we can solve symbolically.
The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
and $\gamma_2$ nonnegative.
If $C_{11} \neq C_{12}$, 
we can write the solution with $\gamma_1$ as the free variable:
\begin{align*}
G_{12} &= \frac{1}{2(C_{11} - C_{12})} - \frac{C_{11}}{2(C_{11} - C_{12})}\gamma_1 \\
G_{21} &= \frac{-2(C_{11} - C_{12}) + (C_{12} - C_{22})}{2(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})}{2(C_{11} - C_{12})^2}\gamma_1 \\
\gamma_2 &= \frac{(C_{11} - 2C_{12} + C_{22})^2}{C_{22}(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})^2}{C_{22}(C_{11} - C_{12})^2}\gamma_1.
\end{align*}
This implies that $\gamma_2$ decreases as $\gamma_1$ increases as long as $C_{12} \neq C_{22}$.
This makes sense because
increasing a rate of coalescence cannot make expected coalescence times longer,
so in order to keep coalescence times the same,
if one coalescence rate is increased, another must be decreased.
This also implies that, in order for $\gamma_2$ to be a value other than $0$, 
we must have $C_{11} - 2C_{12} + C_{22} \neq 0$.
In general, if there is a value or range of values of $\gamma_1$
for which the other movement and coalescence parameters are non-negative,
then a solution exists.

@@

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
the movement rates, $G_{ij}$, that are not constrained to be zero
and the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph,
and only movement rates corresponding to edges in the graph are allowed to be nonzero.
Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
that there are $m_G$ edges in the graph,
and that each of the $n$ populations has its own coalescence rate.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
    \begin{split}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
    \end{split}
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.

Inference with commute times is done in exactly the same way,
except that $q$ replaces $\gamma$,
and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.

@@

We compared inference under coalescence and commute time models
using both data ``from the model'' and more realistic simulations.
For the first category, we designed a number of graphs to test particular aspects of inference,
with migration rates specified on each directed edge
(plotted below using igraph \citep{igraph}).
Given the edge weights ($G$) and the coalescence rates ($\gamma$) of a graph,
we produced data by calculating exact coalescence times ($\mathcal{C}(G,\gamma)$),
to whose entries we added independent Gaussian noise while preserving symmetry,
as described above.
Particular parameters and levels of noise are given in the Results.

@@

We first tested the method using data generated under the model.
To test the methods across a wide range of situations,
we generated random movement rates for 
a large number of $3 \times 2$ rectangular grids:
twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
We generated movement parameters
by rounding independent exponential random variables up to the nearest one tenth,
and fixed coalescence rate at 1 for all locations.
We then added independent noise to each matrix entry (preserving symmetry),
and assume that we can estimate the amount of noise reasonably well, 
so used the standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
To assess the impact on inference of the number of parameters,
we performed inference both with and without the assumption 
that coalescence rate is the same everywhere.
For $3 \times 2$ graphs, there are 21 equations
and 14 movement parameters,
so that with a global coalescence rate there are 15 unknowns
but with individual coalescence rates there are 20 unknowns,
which is close to nonidentifiable.
(We will not present results from an analogous version of resistance distance inference
that constrains all values of $q$ to be identical
because the true values are not identical, even if all coalescence rates are,
and in practice the method infers true values of $q$ extremely reliably.)
To quantify accuracy across these random graphs in different situations,
we computed, for each model fit on a particular random graph,
the absolute difference between the posterior median of each movement parameter
and the true value.
We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
reported below as ``mean absolute error''.

@@

We then test the method's ability to locate barriers to migration
on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
Each nonzero movement rate is determined randomly as before.
Coalescence rates are set to 1 everywhere.
Figure \ref{fig:5x3boxplots} shows posterior distributions of the movement rates
for both coalescence and commute time inference
on this graph, with noise standard deviation equal to
1/1000 of the mean coalescence time (left)
and 1/100 of the mean coalescence time (right).
Although increased noise in the data increased uncertainty in the estimated values,
coalescence time inference correctly inferred not only the location of the barrier
but also the migration rates along all other edges within a reasonable margin of error.
Commute time inference also identified the barrier,
but was much less accurate for the other values.
We also explored making the problem harder,
by first (a) dropping rows and columns from the data corresponding to populations 2, 5, 7, and 15;
and then (b) allowing a separate coalescence rate for each location,
in both cases using a noise level of 1/100.
The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
and show substantially increased error,
but only in movement rates to the removed locations.
As before, uncertainty was increased when multiple coalescence rates were allowed,
but not as strongly as before, likely because the problem is less close to being underconstrained,
having 105 equations and 59 unknowns.

\begin{figure}
\centering
     \includegraphics[scale=.6]{figs/5x3b_grid}
    \caption{Plot of the grid structure 
    for the 5 $\times$ 3 grid with two barriers:
    movement rates between $\{2,3\}$ and $\{7,8\}$ are zero,
    as are those between $\{8,9\}$ and $\{13,14\}$.
    Values for the non-zero movement rates 
    were chosen by rounding independent draws of an exponential random variable with mean 1 
    up to the nearest tenth.
    } \label{fig:5x3b_grid}
\end{figure}

\begin{figure}
\centering
     \includegraphics[scale=0.8]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
    \plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
         put this figure in the supplement}
}
    \label{fig:5x3boxplots}
\end{figure}

@@

Both matrices of coalescence and commute times, $C$ and $R$, are symmetric,
but they do not deal with asymmetric (biased) migration in the same way.
Since commute time to $i$ requires paths both to and from $i$,
but coalescence time to $i$ only reaquires lineages to leave $i$,
if, for instance, there is a low rate of migration back into $i$,
then commute times to $i$ could be quite long even while coalescence times are not.
For instance, suppose there are three populations arranged in a line,
and lineages currently in the outer populations are much more likely to come from the central population
than the other way around.
Since lineages quickly move to the center and coalesce regardless of starting position,
coalescence time (and genetic diversity) will be relatively low between all individuals.
Commute time between the two outer locations, on the other hand,
will be much longer than other comparisons, since it requires a lineage to leave the center.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/4x4coalvcomH}
    \caption{Plots of the true values of $\comdist$ vs $C$ 
    for the uniform graph, symmetric graph, the asymmetric graph,
    and the biased asymmetric graph, 
    shown with the $y=x$ line.
    \plr{TODO: make this a 2x2 grid instead of 4x1}
    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
    \label{fig:4x4coalvcomH}
\end{figure}
